% vim: set foldmethod=marker foldlevel=0:

\documentclass[a4paper]{article}
\usepackage[UKenglish]{babel}

\usepackage{preamble}

\fancyhead[L]{MA150 Assignment 3}
\title{MA150 Algebra 2, Assignment 3}

\begin{document}

\maketitle

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

% {{{ Q6
\question{6}

\begin{equation}
W = (x + 2y - 3z = 0) \subset \R^3
\label{eqn:Q6-W}
\end{equation}

\subsection{~}

The vector $\begin{pmatrix}1\\ 1\\ -1\end{pmatrix}$ is not in $W$ since it doesn't satisfy the equation. In particular, $1(1) + 2(1) - 3(-1) = 6 \ne 0$. Therefore $W \ne \R^3$.

We know from lectures that the dimension of a subspace is less than or equal to the dimension of the parent space, and they have the same dimension if and only if they are equal. Since $W \subset \R^3$, $\dim W \le \dim \R^3$. The dimension of $\R^3$ is 3 (since the standard basis of $\R^3$ has 3 elements). Therefore $\dim W \le 3$. But $W \ne \R^3$, so $\dim W < 3$.

\subsection{~}

We can rearrange equation~(\ref{eqn:Q6-W}) to get $x = 3z - 2y$. Then we can introduce parameters $\lambda$ and $\mu$ and conclude that any point in $W$ can be written as $$\begin{pmatrix}3 \mu - 2 \lambda\\ \lambda\\ \mu\end{pmatrix} = \lambda \begin{pmatrix}-2\\ 1\\ 0\end{pmatrix} + \mu \begin{pmatrix}3\\ 0\\ 1\end{pmatrix}$$
Therefore $\l\{ \begin{pmatrix}- 2\\ 1\\ 0\end{pmatrix}, \begin{pmatrix}3\\ 0\\ 1\end{pmatrix} \r\}$ is a basis of $W$.

Call the elements of this basis $\{ w_1, w_2 \}$ for convenience. Plugging $w_1$ into equation~(\ref{eqn:Q6-W}) gives $1(-2) + 2(1)  - 3(0) = 0$ as required, and plugging $w_2$ into equation~(\ref{eqn:Q6-W}) gives $1(3) + 2(0) - 3(1) = 0$ as required. Therefore $w_1, w_2 \in W$.

For $w_1$ and $w_2$ to be independent, we need to show that $\lambda w_1 + \mu w_2 = 0_W$ if and only if $\lambda = \mu = 0$. That linear independence equation expands to $$\begin{pmatrix}3 \mu - 2 \lambda\\ \lambda\\ \mu\end{pmatrix} = \begin{pmatrix}0\\ 0\\ 0\end{pmatrix}$$
The second components of the vectors imply $\lambda = 0$, and the third components imply $\mu = 0$. Therefore $w_1$ and $w_2$ are linearly independent.

$w_1$ and $w_2$ must span $W$ since any linear combination is of the form $\begin{pmatrix}3 \mu - 2 \lambda\\ \lambda\\ \mu\end{pmatrix}$ and we showed before that that is equivalent to equation~(\ref{eqn:Q6-W}), which is the definition of $W$.

Since we have a basis of $W$ with 2 elements, we know that $\dim W = 2$.

% }}}

% {{{ Q7
\newquestion{7}

$$\varphi(1) = \begin{pmatrix}1\\ 0\end{pmatrix}, \quad \varphi(x) = \begin{pmatrix}-1\\ 1\end{pmatrix}, \quad \varphi(x^2) = \begin{pmatrix}1\\ -2\end{pmatrix}, \quad \varphi(x^3) = \begin{pmatrix}-1\\ 3\end{pmatrix}$$

\subsection{~}

\begin{align*}
\varphi(2x^3 - 3x + 2) &= \varphi(2x^3) + \varphi(-3x) + \varphi(2)\\[1ex]
&= 2 \varphi(x^3) - 3 \varphi(x) + 2 \varphi(1)\\[1ex]
&= 2 \begin{pmatrix}-1\\ 3\end{pmatrix} - 3 \begin{pmatrix}-1\\ 1\end{pmatrix} + 2 \begin{pmatrix}1\\ 0\end{pmatrix}\\[1ex]
&= \begin{pmatrix}-2\\ 6\end{pmatrix} + \begin{pmatrix}3\\ -3\end{pmatrix} + \begin{pmatrix}2\\ 0\end{pmatrix}\\[1ex]
&= \begin{pmatrix}3\\ 3\end{pmatrix}
\end{align*}

\subsection{~}

$$\psi(f) = \begin{pmatrix} f(-1) \\[1ex] \ds \dd fx (-1) \end{pmatrix}$$

By proposition 5.17, two linear maps are equal if their domains and codomains are equal and they agree on the elements of a basis of the domain. $\varphi$ and $\psi$ are both defined on $\varphi, \psi : \R[x]_{\le 3} \to \R^2$. Then we just have to check that $\varphi$ and $\psi$ agree on some basis of the domain, and it makes sense to use $\{ 1, x, x^2, x^3 \}$.

\begin{gather*}
\psi(1) = \begin{pmatrix}1\\ 0\end{pmatrix},\quad
\psi(x) = \begin{pmatrix}-1\\ 1\end{pmatrix},\quad\\[1ex]
\psi(x^2) = \begin{pmatrix}(-1)^2\\ 2(-1)\end{pmatrix} = \begin{pmatrix}1\\ -2\end{pmatrix},\quad
\psi(x^3) = \begin{pmatrix}(-1)^3\\ 3(-1)^2\end{pmatrix} = \begin{pmatrix}-1\\ 3\end{pmatrix}
\end{gather*}

Since $\psi$ and $\varphi$ agree on a basis, $\psi = \varphi$.

\subsection{~}

To find the image of a linear transformation, we can write it as a matrix and take the column span of its row reduced echelon form. $\varphi$ is $L_M$ where $$M = \begin{pmatrix} 1 & -1 & 1 & -1\\ 0 & 1 & -2 & 3 \end{pmatrix}$$

Finding $\rref(M)$ only takes one step, $A_{21}(1)$. $$\rref(M) = \begin{pmatrix} 1 & 0 & -1 & 2\\ 0 & 1 & -2 & 3 \end{pmatrix}$$
Then $\op{Colspan}(\rref(M)) = \l\{ \begin{pmatrix}1\\ 0\end{pmatrix}, \begin{pmatrix}0\\ 1\end{pmatrix} \r\}$, so $\Im \varphi = \R^2$.

\subsection{~}

$\varphi$ is defined on the domain $V = \R[x]_{\le 3}$, which has dimension 4. Also $\Im \varphi = \R^2$, so $\dim \Im \varphi = 2$. Therefore by the Rank-Nullity Theorem, $$\dim \ker \varphi = \dim V - \dim \Im \varphi = 4 - 2 = 2$$

% }}}

% {{{ Q8
\newquestion{8}

$$\pi(x^i) = (x+a)^i$$

\subsection{~}

An isomorphism of vector spaces is just a bijective linear map. We shall first prove that $\pi$ is a linear map.

We expect $\pi(\lambda x^i) = \lambda \pi(x^i)$. \begin{align*}
\pi(\lambda x^i) &= \pi\l( \l( \lambda^{\f1i} x \r)^i \r)\\[1ex]
&= \l( \lambda^{\f1i} x + a \r)^i\\[1ex]
\end{align*}

% TODO: This question sucks and I don't understand it so I'm giving up

\subsection{~}

$$L_\pi = \begin{pmatrix} 1 & a & a^2\\ 0 & 1 & 2a \\ 0 & 0 & 1 \end{pmatrix}$$

% }}}

% {{{ Q9
\newquestion{9}

$$V = \l\{ f : \R \to \R : f \text{ is differentiable} \r\}, \quad W = \l\langle \cos(x), \cos(2x) \r\rangle$$

\subsection{~}

$\cos(x)$ and $\cos(2x)$ are linearly independent and span $W$ by definition, so $\{\cos(x), \cos(2x)\}$ is a basis for $W$. The dimension of a vector space is equal to the number of vectors in a basis, so $\dim W = 2$.

\subsection{~}

Let $\cal U = \{ f \in W : f(10) = 0 \}$. So we want functions of the form $$\lambda \cos(x) + \mu \cos(2x)$$
for some $\lambda, \mu \in \R$ where $\lambda \cos(10) + \mu \cos(20) = 0$. That means we need $$\lambda = \f{-\mu \cos(20)}{\cos(10)}$$

Therefore every element of $\cal U$ is of the form $$\mu \l(\f{-\cos(20)}{\cos(10)} \cos(x) + \cos(2x) \r)$$
and therefore $\l\{ \df{-\cos(20)}{\cos(10)} \cos(x) + \cos(2x) \r\}$ is a basis of $\cal U$. Since this basis has 1 element, $\dim \cal U = 1$.

\subsection{~}

Let $\cal U_2 = \{ f \in W : f(10) = 1 \}$. So we want functions of the form $\lambda \cos(x) + \mu \cos(2x)$ for some $\lambda, \mu \in \R$ where $\lambda \cos(10) + \mu \cos(20) = 1$. Is $\cal U_2$ a subspace of $W$?

To be a subspace of $W$, $\cal U_2$ must be a non-empty subset (this is trivially true), and must be closed under the operations of $W$. So if we have some $\alpha \cos(x) + \beta \cos(2x) \in \cal U_2$, then we want $\lambda \l( \alpha \cos(x) + \beta \cos(2x) \r) \in \cal U_2$ for any $\lambda \in \R$.

But $\alpha \cos(10) + \beta \cos(20) = 1$ by definition of $\cal U_2$, and $\lambda \l( \alpha \cos(x) + \beta \cos(2x) \r) = \lambda \ne 1$. Therefore $\cal U_2$ is not closed under scalar multiplication and therefore is not a subspace of $W$.

% }}}

% {{{ Q10
\newquestion{10}

$$L_A : \R^4 \to \R^3, \quad A = \begin{pmatrix} 1 & 2 & 3 & 4 \\ 0 & 0 & a & b \\ 0 & 0 & c & d \end{pmatrix}$$

\subsection{~}

Let $\varphi : V \to W$ be a linear map. Then $\dim \Im \varphi + \dim \ker \varphi = \dim V$.

\subsection{~}

$a = b = c = d = 0$ gives $\dim \op{Colspan} A = 1$. Then $$\dim \Im L_A = \dim \op{Colspan} A = 1$$
and then by the Rank-Nullity Theorem, $\dim \ker L_A = 4 - 1 = 3$.

\subsection{~}

$a = 1$, $b = c = d = 0$ gives $\dim \op{Colspan} A = 2$. Then $\dim \Im L_A = 2$ and $\dim \ker L_A = 2$.

\subsection{~}

$a = 1$, $d = 1$, $b = c = 0$ gives $\dim \op{Colspan} A = 3$. Then $\dim \Im L_A = 3$ and $\dim \ker L_A = 1$.

% }}}

\end{document}
