% vim: set foldmethod=marker foldlevel=0:

\documentclass[a4paper]{article}
\usepackage[UKenglish]{babel}

\usepackage{preamble}
\usepackage{tikz}

\title{CS147 Discrete Maths and its Applications 2, Assignment 2}
\author{Dyson Dyson (ID: 5503449)}
\date{}

\begin{document}

\maketitle

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

% {{{ Q1
\question{1}

Let $G = (V, E)$ be a graph, $M \subset E$ be a matching on $G$, and $Z \subset M$.

A matching is just a set of edges in $G$ which do not share any common endpoint. For $Z$ to not be a matching, we would need to choose two edges from $M$ which share a node. Since $M$ is a matching, no such pair of edges exists by definition, so $Z$ must also be a matching.

% }}}

% {{{ Q2
\question{2}

Let $G = (V, E)$ be a graph with $n$ nodes, where each node $v \in V$ is incident on exactly 2 edges.
%By Euler's handshaking lemma, there must be $n$ edges.
The graph is isomorphic to an $n$-gon. For example, the case of $n=4$ could be drawn as a square.

In the case of even $n$, there must exist two maximum matchings of size $\df n2$, which are complements of each other in $E$, so $M_2 = E \setminus M_1$.

In the case of odd $n$, we still get two complementary matchings, but they are not large enough. Take the case of $n=5$ for example,

\begin{center}
\begin{tikzpicture}[
	every path/.style={thick},
]
	\node (a) at (90:1) {$a$};
	\node (b) at (162:1) {$b$};
	\node (c) at (234:1) {$c$};
	\node (d) at (306:1) {$d$};
	\node (e) at (18:1) {$e$};

	\draw [red] (a) -- (b);
	\draw (b) -- (c);
	\draw [red] (c) -- (d);
	\draw (d) -- (e);
	\draw (e) -- (a);

	\node (a) at ([xshift=3cm]90:1) {$a$};
	\node (b) at ([xshift=3cm]162:1) {$b$};
	\node (c) at ([xshift=3cm]234:1) {$c$};
	\node (d) at ([xshift=3cm]306:1) {$d$};
	\node (e) at ([xshift=3cm]18:1) {$e$};
                        
	\draw (a) -- (b);   
	\draw [red] (b) -- (c);
	\draw (c) -- (d);
	\draw (d) -- (e);
	\draw [red] (e) -- (a);
\end{tikzpicture}
\end{center}

Both of the subsets highlighted in red are matchings, but both are maximum and of size 2. It is clear that in the case of odd n, a maximum matching has size $\l\lfloor \df n2 \r\rfloor$.

Therefore it is false that there must exist a matching $M \subset E$ with $|M| \ge \df n2$.

% }}}

% {{{ Q3
\question{3}

Let $G = (V, E)$ be a graph, and $M$ be a maximal matching on $G$, so every edge in $E \setminus M$ has at least one endpoint that is matched under $M$, meaning $M$ cannot be extended. Also let $M^*$ be a matching of maximum size on $G$.

Is it true that $|M| \ge \f12 |M^*|$? Yes.

Suppose we have a situation where $|M| < \f12 |M^*|$. Let $|M| = \ell$ and $|M^*| = k$ so that $M$ matches $2\ell$ nodes and $M^*$ matches $2k$ nodes. The inequality implies $\ell < \f12 k \iff 2\ell < k$.

There are at most $2\ell$ edges in $M^*$ which are matched by $M$. But since $2\ell < k$, there is at least one edge in $M^*$ which is not matched by $M$. Therefore we can add this edge to $M$, meaning it is not maximal. That's a contradiction, therefore $|M| < \f12 |M^*|$.

% }}}

% {{{ Q4
\question{4}

Let $G = (L \cup R, E)$ be a bipartite graph where every edge connects one node in $L$ to one node in $R$, and let $N_G(X)$ denote the neighbours of some set $X$ of nodes in $G$. Also $G$ has the property that for all $A \subset L$, $|N_G(A)| \ge \f12 |A|$.

We want to know if there exists a subset $H \subset E$ where $|H| = |L|$, every node in $L$ is incident upon exactly one edge in $H$, and every node in $R$ is incident upon at most two edges in $H$.

To satisfy the first two properties, we require that $H$ is constructed by considering each node in $L$ and choosing one of the edges that connects to it.

Is it possible that there exists a $v \in R$ which is incident on three edges in $H$? Suppose such a $v$ does exist. Then those three edges in $H$ would connect to three distinct nodes in $L$, call them $S = \{u_1, u_2, u_3\}$. But by the neighbour requirement of $G$, we have $|N_G(S)| \ge \f12 |S|$.

By construction, all three nodes connect to the same $v \in R$ and no other nodes, so $N_G(S) = 1$. Therefore we have $1 \ge \f32$, which is a contradiction.

Therefore we cannot have a $v \in R$ which is incident on three edges in $H$. Therefore every node in $R$ is incident on at most two edges in $H$, so the statement is true.

% }}}

\end{document}
